WSClean: an implementation of a fast, generic wide-field imager for radio astronomy

### 2014-07-01 articles/2014/Article-WSClean/Response-to-referee.txt

I have asked the director of the MWA board, S. Tingay, to respond to the editor's question on the appropriateness of the large number of coauthors. He asked me to forward the following answer:

"The authorship of this paper, and other similar papers published by the MWA consortium, contain a significant number of authors.  The authorship consists of people who have made a significant intellectual contribution to the establishment of the MWA (development of the instrumentation; deployment, testing, and commissioning of the instrument) as well as people who have undertaken the substantial observations and data analysis, and the interpretation of the data (in the case of this paper, as processed using the WSClean software.  The authors are people either funded to contribute to the MWA or are people who work on the MWA project in the research allocations of their positions.  All contributions are meaningful and are audited by the MWA Director and Board.  These contributions should be justly recognised in publications such as this."

We have marked the first author as a corresponding author, as suggested by the editor.

Additionally, we had taken off one co-author and added co-author B. Pindor, who had accidentially been left off.

We would like to thank the reviewer for his thorough review. We agree on almost all his comments, have changed the paper accordingly, and we think that his improved the clarity of the paper. We will provide detailed answers to his comments below. We have marked all changes with bold and red in the revised article.

p. 2 col. 1: “To reach high dynamic ranges, it can be necessary to deal with DDEs…” As worded, the paragraph seems to imply that A-projection is the complete answer to the DDE problem. A-projection will only correct for DDEs that are known. This can only be done directly for deterministic effects like primary beam rotation; effects like the ionosphere require separate calibration or estimation before A-projection can be applied.

Answer: We've added a sentence to clarify this.

p.2 col. 2: “”which we believe to be the fastest generic wide-field imager” such statements are good for talks, but not for papers. At least two GPU-based imagers purport to be faster, at which point one starts splitting hairs about terms like “generic” and “available”. I’d be more comfortable with a statement like “Our experience suggests that WSClean is anywhere from 2 to 12 times faster than mainstream w-projection imagers (CASA, lwimager) on the same datasets.” This is already impressive enough!

Answer: we have changed this to: "Our new imaging implementation, which in our experience is anywhere from 2 to 12 times faster than the CASA w-projection imager, is publicly released\footnote..."

Section 2: this needs a reference to Cornwell, or whoever suggested the w-stacking approach first.

Answer: We've added a citation to "Analysis of Convolutional Resampling Algorithm Performance" by B. Humphreys and T. J. Cornwell.

p. 3 col. 2: “Using too few w-layers…” The term w-aliasing is introduced unexplained (or is the paragraph meant to be the explanation?) In any case a better explanation of the effect is needed.

Answer: We've reworded this.

p. 4 Sect 3.2. This section needs more clarity, especially the last paragraph of p. 4. There needs to be a clear comparison between the Cornwell et al. 2012 definition of w-snapshots, and the method proposed here. I recommend splitting Sect. 3.2 in two, and moving the theoretical description of w-snapshots to Section 2, while leaving implementation details in Section 3.

Answer: We have moved the theoretical description to a new subsection in Section 2 as suggested, and added some text to explain the method more clearly.

Section 3 in general: recommend clearly splitting into generic implementation details, and MWA-specific details, as it is currently not clear how much is MWA-specific and whether this limits functionality in any way.

Answer: We have added a sentence to Sect 3.2 to explain that the method is generic, and renamed Sect 3.3 to include "for the MWA".

p. 7: “Our image-based beam-correction…” This can only work if the beam is also constant in time. How is sky rotation accounted for?

Answer: we account for this by doing snapshot imaging, and assuming a constant beam over the snapshot interval. We have reworded this sentence to say: "Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but requires snapshot imaging, because the MWA beam changes over time because of Earth rotation, and only works because all tiles have the same beam."

p.7. “CASA convolves samples” better “CASA’s gridder implementation convolve samples”.

Answer: Reworded.

p.7 Sect 3.5.: “Decreasing inversion resolution” quite a mouthful, suggest revising title. “Lowering the resolution of reverse transforms”, perhaps?

Answer: Renamed to "Lowering the resolution of inversion"

p. 7 Figure 5: This figure needs labels, including axis labels.

Answer: We reworded the axis labels and reworded the figure caption.

p. 7 Sect. 3.5.: suggest use of “PSF” rather than “synthesized-beam” to avoid confusion with primary beam.

Answer: Not changed, we think that 'synthesized beam' should be a known term for readers of the article, and we think that talking about 'synthesized-beam width' makes a bit more sense than talking about 'PSF width'.

p. 7 Sect 3.5: “Therefore inversion at high resolution is not required…” and until end of paragraph, please improve clarity.

Answer: We have reworded this paragraph.

p. 7 “The outputs with and without rescaling”. The term “rescaling” is dropped in unannounced, does this simply refer to the resolution-lowering optimization described in this section? If so, this should be clearly defined.

Answer: Changed to: "The outputs with and without lowering the inversion and prediction resolution do not visibly differ, but ..."

p. 8 end of Sect 3.5: RMS comparison per se is somewhat meaningless without referencing the source flux (i.e. effectively the DR of the image). Do I interpret the colour bars of Figure 3 correctly, that this is a 10:1 DR image? Have the authors explored whether/how the lowered resolution affects things in mid- to high-DR situations? Given the nature of the optimization, I would expect the errors to scale with model flux, i.e. to be more noticeable at high DR. If so, this is an important caveat that should be noted.

Answer: We have added the peak flux level in the image (10 Jy) and clarified that the quoted noise level was in the residual image. If DR is defined as peak flux / noise level, then this image is a 1:150 DR image. In Section 4.1, a 1:1000 simulated scenario with a simpler model is tested, and as described there, no difference is visible between images with and without the method.

Figure 7: What does R-WSCLEAN refer to? I assume “recentering”, but this should be made clear. Or is it “rescaling”? If the former, and since “recentering” seems to be equivalent to w-snapshots (as Sect 3.2. implies, but could make clearer), perhaps this should be made clear in the figure caption.

Answer: Clarified in the caption; it refers indeed to recentring.

p.10 col. 2: “Imaging the MWA primary beam requires 3072 pixels… decreases to about 1500…” I thought the low-res optimization does not apply to the clean model image per se, but only kicks in during the reverse transform? if so, this sentence is confusing.

Answer: Reworded to: "If the small-inversion optimisation of \S3.5 is used, the inversion can be performed at an image size of 1500 pixels."

p. 12 Table 5: If R-WSCLEAN is equivalent to w-snapshot, why are we seeing almost exactly the same performance figures for w-stacking and w-snapshot? This seems to contradict the curves of Figure 7.

Answer: This is because Fig. 7 tests larger image sizes with more costly scenarios, which makes the w-snapshot is more effective. However, with the scenarios from Table 4, there is no benefit. We have added the following explanation to clarify this:
After "The w-snapshot method does not improve the performance much over normal WSCLEAN operation in any of the cases."
We added: "This might seem to differ from some of the results in Fig. 7, where the w-snapshot method does show improvement in certain cases. This is because Fig. 7 tests somewhat more extreme parameters, in which the w-snapshot method shows more benefit. The w-snapshot might become more valuable at higher ZAs or when the image size needs to be larger than the half-power beam width."

p. 12 Table 5: is the lowered-resolution optimization on for these performance comparisons? This should be stated. If yes, then the DR should be stated -- the caveat is, do we know how the optimization behaves at high DRs?

Answer: We have clarified in 4.6 that this assumes the optimization. We have added to Sect 5 the DR of the simulated observation (1:1000).

p. 12 col. 1: “The image size is set to two times the FOV width divided by the resolution”. Is this justified, given that the authors discuss the need for x5 oversampling for CLEAN earlier on?

Answer: This is justified by the fact that the small-inversion size optimization can be used, which we have no clarified.

p. 14 col. 1: “Our optimisation of lowering the image resolution … has no noticeable effect…” See remarks above, has this been tested at high DR?

Answer: Clarified, see earlier answer.

p. 14 col. 1: “It was suggested that w-snapshots would improve…” I do not understand the sentence, please rephrase.

Answer: Reworded.

### 2014-07-03 articles/2014/Article-WSClean/wsclean.tex
Removing editmarks

\newcommand{\editmark}[1]{#1}
%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$,
R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, 
J.~Rhee\ANU$^,$\CAASTRO, \newauthor
T.~Murphy\USydney$^,$\CAASTRO,
N.~D.~R.~Bhat\Curtin,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
A.~Ewall-Wice\MIT,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT, \newauthor
L.~J.~Greenhill\CfA, 
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
M.~Johnston-Hollitt\Victoria, \newauthor
D.~C.~Jacobs\ASU,
J.~C.~Kasper\UMichigan$^,$\CfA,
E.~Kratzenberg\Haystack, 
E.~Lenc\USydney$^,$\CAASTRO,
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin, \newauthor
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO,
M.~F.~Morales\UW, 
E.~Morgan\MIT,
N.~Kudryavtseva\Curtin, \newauthor
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
\editmark{B.~Pindor\UMelbourne,}
P.~Procopio\UMelbourne,
T.~Prabu\RRI,
J.~Riding\UMelbourne,  \newauthor
%A.~E.~E.~Rogers\Haystack,
D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO,
S.~J.~Tingay\Curtin$^,$\CAASTRO, \newauthor
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~R.~Whitney\Haystack, 
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
}

@@

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for wide-field telescopes. One way to \editmark{correct for known} DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}. One particular DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all antennas, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \textsc{awimager} \citep{awimager-2013} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking. \editmark{The $a$-projection technique can only be applied directly for deterministic effects such as the correction of the primary beam. Effects like the ionosphere require separate calibration or estimation before $a$-projection can be applied.}

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{casa}'s $w$-projection implementation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which \editmark{in our experience is anywhere from 2 to 12 times faster than the CASA $w$-projection imager}, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.

@@

\section{The $w$-stacking technique} \label{sec:wstacking}
In this section, we will describe the $w$-stacking algorithm from a mathematical point of view. \editmark{Instead of applying a convolution in $uv$-space, the $w$-stacking method grids visibilities on different $w$-layers and performs the $w$-corrections after the inverse Fourier transforms \citep{ska-memo-regridding-2011}.}

@@

\editmark{To grid the visibilities, the $w$-values are rounded to the $w$-value of the nearest $w$-layer. This discretisation can cause noticeable aliasing when using too few $w$-layers, and results in decorrelation of sources far from the phase centre in the longer baselines. Additionally, this $w$-aliasing can cause ghost sources to appear} in the image. An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}. When Cotton-Schwab cleaning includes prediction with too few $w$-layers, incorrect values will be subtracted from the visibilities even when no aliased sources are cleaned during minor iterations. Therefore, accurate prediction is more important than accurate imaging, because aliasing artefacts are attenuated by cleaning as long as the model is subtracted accurately.

@@

\subsection{\editmark{$W$-snapshot imaging}} \label{sec:snapshot-imaging-theory}
\editmark{$W$-snapshot imaging is a technique that combines warped-snapshot imaging with a $w$-correcting technique, such as $w$-projection or $w$-stacking \citep{widefield-imaging-ska-cornwell}. In the warped-snapshot imaging technique, the $w$-term is neglected, which results in an image with distorted coordinates \citep{perley-noncoplanar-arrays,mwa-interferometric-imaging}. Additionally, if the positions of the array elements are not perfectly planar or multiple timesteps are integrated to the same grid, visibilities will decorrelate and this can cause imaging artefacts. The original $w$-snapshot algorithm as described in \citet{widefield-imaging-ska-cornwell} corrects such artefacts by gridding visibilities on a tilted best-fit plane in $uvw$-space, and performs $w$-corrections towards that plane using $w$-projection or $w$-stacking.}

\editmark{We have looked into implementing snapshot imaging with $w$-corrections in a slightly different way. Instead of performing transforms of tilted planes, our method consists of phase-rotating the visibilities such that the phase centre of the observation is towards the zenith direction, i.e., the direction with minimal $w$-values. During imaging, the $w$-layers are recentred from zenith to the direction of interest by phase-shifting the visibilities, thereby translating the image over the tangent plane. This method results in $w$-corrected snapshots which need to be regridded before further integration, similar to the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, this method transforms planes with constant $w$-values and produces images in zenith projection.}

\editmark{An} image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentring an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. \editmark{By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.}

\editmark{Our main reason for developing this method is that it is easier to implement, because no changes are required to the performance-critical gridding step. Another} benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning. Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}. A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre. With our approach, the effect gets worse with the distance from zenith. Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.

@@

The new imager, called ``\textsc{wsclean}'', is written in the C++ language. It reads visibilities from \textsc{casa} measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimisations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimisation \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced point-spread function (PSF), less relevant. For very large images this optimisation might still be useful, but we have not implemented it in \textsc{wsclean}. Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}. \textsc{wsclean} allows cleaning individual polarisations, or can jointly deconvolve the polarisations. \editmark{In the latter case, peak finding can be performed in the sum of squared Stokes parameters, $I^2+Q^2+U^2+V^2$, or in} $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$. After a peak has been found, the PSF is subtracted from the individual polarisations with different factors.

@@

For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because they are scientifically of interest (e.g. when searching for transients). Full-sky imaging can also be useful for self-calibration or deconvolution. For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., \editmark{submitted}). An example of a full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.

@@

\editmark{Our implementation supports cleaning of recentred images in the same modes that normal images can be cleaned.} The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{\editmark{Gridding kernel size plotted against imaging time, using common MWA settings.}}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}

>>>>

%\newcommand{\editmark}[1]{#1}
\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{Corresponding author. E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$,
R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, 
J.~Rhee\ANU$^,$\CAASTRO, \newauthor
T.~Murphy\USydney$^,$\CAASTRO,
N.~D.~R.~Bhat\Curtin,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
A.~Ewall-Wice\MIT,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT, \newauthor
L.~J.~Greenhill\CfA, 
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
M.~Johnston-Hollitt\Victoria, \newauthor
D.~C.~Jacobs\ASU,
J.~C.~Kasper\UMichigan$^,$\CfA,
E.~Kratzenberg\Haystack, 
E.~Lenc\USydney$^,$\CAASTRO,
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin, \newauthor
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO,
M.~F.~Morales\UW, 
E.~Morgan\MIT,
N.~Kudryavtseva\Curtin, \newauthor
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
B.~Pindor\UMelbourne,
P.~Procopio\UMelbourne,
T.~Prabu\RRI,
J.~Riding\UMelbourne,  \newauthor
%A.~E.~E.~Rogers\Haystack,
D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO,
S.~J.~Tingay\Curtin$^,$\CAASTRO, \newauthor
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~R.~Whitney\Haystack, 
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
}

@@

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for wide-field telescopes. One way to correct for known DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}. One particular DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all antennas, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \textsc{awimager} \citep{awimager-2013} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking. The $a$-projection technique can only be applied directly for deterministic effects such as the correction of the primary beam. Effects like the ionosphere require separate calibration or estimation before $a$-projection can be applied.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{casa}'s $w$-projection implementation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which in our experience is anywhere from 2 to 12 times faster than the CASA $w$-projection imager, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.

@@

\section{The $w$-stacking technique} \label{sec:wstacking}
In this section, we will describe the $w$-stacking algorithm from a mathematical point of view. Instead of applying a convolution in $uv$-space, the $w$-stacking method grids visibilities on different $w$-layers and performs the $w$-corrections after the inverse Fourier transforms \citep{ska-memo-regridding-2011}.

@@

To grid the visibilities, the $w$-values are rounded to the $w$-value of the nearest $w$-layer. This discretisation can cause noticeable aliasing when using too few $w$-layers, and results in decorrelation of sources far from the phase centre in the longer baselines. Additionally, this $w$-aliasing can cause ghost sources to appear in the image. An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}. When Cotton-Schwab cleaning includes prediction with too few $w$-layers, incorrect values will be subtracted from the visibilities even when no aliased sources are cleaned during minor iterations. Therefore, accurate prediction is more important than accurate imaging, because aliasing artefacts are attenuated by cleaning as long as the model is subtracted accurately.

@@

\subsection{$W$-snapshot imaging} \label{sec:snapshot-imaging-theory}
$W$-snapshot imaging is a technique that combines warped-snapshot imaging with a $w$-correcting technique, such as $w$-projection or $w$-stacking \citep{widefield-imaging-ska-cornwell}. In the warped-snapshot imaging technique, the $w$-term is neglected, which results in an image with distorted coordinates \citep{perley-noncoplanar-arrays,mwa-interferometric-imaging}. Additionally, if the positions of the array elements are not perfectly planar or multiple timesteps are integrated to the same grid, visibilities will decorrelate and this can cause imaging artefacts. The original $w$-snapshot algorithm as described in \citet{widefield-imaging-ska-cornwell} corrects such artefacts by gridding visibilities on a tilted best-fit plane in $uvw$-space, and performs $w$-corrections towards that plane using $w$-projection or $w$-stacking.

We have looked into implementing snapshot imaging with $w$-corrections in a slightly different way. Instead of performing transforms of tilted planes, our method consists of phase-rotating the visibilities such that the phase centre of the observation is towards the zenith direction, i.e., the direction with minimal $w$-values. During imaging, the $w$-layers are recentred from zenith to the direction of interest by phase-shifting the visibilities, thereby translating the image over the tangent plane. This method results in $w$-corrected snapshots which need to be regridded before further integration, similar to the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, this method transforms planes with constant $w$-values and produces images in zenith projection.

An image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentring an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.

Our main reason for developing this method is that it is easier to implement, because no changes are required to the performance-critical gridding step. Another benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning. Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}. A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre. With our approach, the effect gets worse with the distance from zenith. Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.

@@

The new imager, called ``\textsc{wsclean}'', is written in the C++ language. It reads visibilities from \textsc{casa} measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimisations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimisation \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced point-spread function (PSF), less relevant. For very large images this optimisation might still be useful, but we have not implemented it in \textsc{wsclean}. Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}. \textsc{wsclean} allows cleaning individual polarisations, or can jointly deconvolve the polarisations. In the latter case, peak finding can be performed in the sum of squared Stokes parameters, $I^2+Q^2+U^2+V^2$, or in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$. After a peak has been found, the PSF is subtracted from the individual polarisations with different factors.

@@

For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because they are scientifically of interest (e.g. when searching for transients). Full-sky imaging can also be useful for self-calibration or deconvolution. For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., submitted). An example of a full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.

@@

Our implementation supports cleaning of recentred images in the same modes that normal images can be cleaned. The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{Gridding kernel size plotted against imaging time, using common MWA settings.}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}
